A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-06-09 analysis/paper/paper.Rmd
Had to correct for a mix-up with the dates from Gunnarsrød 5

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data seperate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data seperate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.
